Understanding the Impact of Mortgage State and Income on Household Poverty Status

### 2024-04-17 paper/paper.qmd
based on TA review

\newpage
# Introduction
Over the last few decades, despite many policies the United States have implemented to mitigate poverty, there has been no real improvement [@Kearney_Harris]. Poverty remains a major issue in the US, affecting millions of people. In 2019, the poverty rate was 10.5% [@Semega_2022]. Therefore, it is important to find out what are the causes of poverty, in order to reduce it. Multiple research, using the Official Poverty Measure, was conducted to look at why poverty persists. However, the Official Poverty Measure is flawed in various aspects, such as not accounting for regional differences in costs of living and government benefits  [@Gesumaria_2015]. So we will analyze why some US household are in poverty in 2019 using the 2019 Revised Supplemental Poverty Measure (SPM) Research dataset from the United States Census Bureau (USCB) [@Bureau_2022]. USCB is an agency of the U.S. Federal Statistical System, responsible for producing data about the American people and economy [@uscb]. Supplemental Poverty Measure was developed to address numerous concerns of Official Poverty Measure critics, providing an improved statistical picture of poverty [@Gesumaria_2015]. We build a prediction model for households poverty status using 2019 Revised SPM dataset. We primarily look at two factors: mortgage status and income in each household. This allows us to look at the question whether potential financial factors may cause the poverty status for US households in 2019.

@@

First, a new column "income after tax" is created by summing up all incomes, credits, and benefits and subtracting all expenses. Then the dataset is cleaned by renaming of column headers, defining column classes, grouping the income variable into different income levels (here the assumed facts are that the income unit is in USD and 'k' stands for thousand, eg. 5k = 10,000), and replacing numerical values in the table with their corresponding descriptions from the data dictionary to improve the readability.

@@

### Model justification
Regarding the relationship between mortgage state and poverty state, we anticipate that both owners without mortgages and owners with mortgages are less likely to experience poverty. Owners without mortgages typically indicate financial stability, as they have already paid off their homes and can allocate financial resources toward household necessities, other than paying back the mortgage / rent. Also, without monthly mortgage or rent payments, their financial burden is reduced, further decreasing the likelihood of poverty. Similarly, owners with mortgages demonstrate financial responsibility by securing loans from banks. The approval of these loans suggests that the bank is confident in owner's ability to manage repayments, indicating a lower possibility of poverty. On the other hand, renters are more likely to experience poverty due to their lack of property ownership and the hidden implication that they do not earn enough to purchase their own property. Renters typically have fewer savings and may earn insufficient salaries to cover rental and essential expenses. Consequently, the combination of rental payments and other financial obligations increases their vulnerability to poverty.

@@

## Relationship betwen Mortgage State and Poverty Status
In the USA in 2019, housing costs accounted for more than 30% of household expenditure, making mortgage status a critical factor in determining a family's risk of falling into poverty [@Statistics_2020]. With 11.8% of the total population (38.3 million people) living below the poverty line, it's evident that housing affordability significantly impacts economic stability. The percentage of individuals below the poverty threshold, adjusted for housing costs are: $29,080 for a family of four with a mortgage, $24,413 for a family of four without a mortgage, and $29,194 for a family of four [@Khattar_Pathak_Schweitzer_Khan_McManus_Jones_Benson_Kelly_2023]. From this, we can assumed that families without mortgages experienced the lowest housing costs, leaving more funds available for other essentials and eliminating the need for alternative housing options. In contrast, families with mortgages may face burdensome monthly payments, especially if they have chosen a floating interest rate, making them vulnerable to fluctuations in interest rates. Renters also encounter challenges, as rent increases can strain finances and necessitate either accepting higher housing costs or undertaking the time-consuming and costly process of finding new accommodations. These housing-related challenges exacerbate the risk of poverty for affected families, underscoring the importance of addressing housing affordability as a key component of poverty alleviation efforts.

## Relationship betwen Income and Poverty Status
The relationship between a household's income after tax and its poverty status is characterized by a negative correlation, wherein higher income corresponds to a reduced probability of experiencing poverty. This correlation shows the role of financial resources in mitigating economic hardship and facilitating socio-economic well-being. However, in the United States, the cost of living is notably high, low-income households often struggle to meet basic needs, resulting in an unending cycle of poverty. In 2019, the poverty guideline for a one-person household was set at $12,490 annually [@ASPE]. So, it is evident that families earning less than $5,000 are definitely living below the poverty line. Conversely, individuals with high incomes are less likely to experience poverty, as their financial resources enable them to comfortably cover living expenses. They will be able to have savings accumulation, further consolidating their financial security and contributing to wealth accumulation over time.

>>>>

\newpage
# Introduction
Despite many policies the United States have implemented to mitigate poverty over the past decades, there has been no real improvement [@Kearney_Harris]. Poverty remains a major issue in the US, affecting millions of people. In 2019, the poverty rate was 10.5% [@Semega_2022]. Therefore, it is important to find out the causes of poverty, in order to reduce it. Multiple research, using the Official Poverty Measure, was conducted to look at why poverty persists. However, the Official Poverty Measure is flawed in various aspects, such as not accounting for regional differences in costs of living and government benefits  [@Gesumaria_2015]. So we will analyze why some US household are in poverty in 2019 using the 2019 Revised Supplemental Poverty Measure (SPM) Research dataset from the United States Census Bureau (USCB) [@Bureau_2022]. USCB is an agency of the U.S. Federal Statistical System, responsible for producing data about the American people and economy [@uscb]. Supplemental Poverty Measure was developed to address numerous concerns of Official Poverty Measure critics, providing an improved statistical picture of poverty [@Gesumaria_2015]. We build a prediction model for households poverty status using 2019 Revised SPM dataset. We primarily look at two factors: mortgage status and income in each household. This allows us to look at the question whether potential financial factors may influence the poverty status for US households in 2019.

@@

First, a new column "income after tax" is created by summing up all incomes, credits, and benefits and subtracting all expenses. Then the dataset is cleaned by renaming of column headers, defining column classes, grouping the income variable into different income levels (here the assumed facts are that the income unit is in USD and 'k' stands for thousand, eg. 5k = 5000), and replacing numerical values in the table with their corresponding descriptions from the data dictionary to improve the readability.

@@

## Model justification
Regarding the relationship between mortgage state and poverty state, we anticipate that both owners without mortgages and owners with mortgages are less likely to experience poverty. Owners without mortgages typically indicate financial stability, as they have already paid off their homes and can allocate financial resources toward household necessities, other than paying back the mortgage / rent. Also, without monthly mortgage or rent payments, their financial burden is reduced, further decreasing the likelihood of poverty. Similarly, owners with mortgages demonstrate financial responsibility by securing loans from banks. The approval of these loans suggests that the bank is confident in owner's ability to manage repayments, indicating a lower possibility of poverty. On the other hand, renters are more likely to experience poverty due to their lack of property ownership and the hidden implication that they do not earn enough to purchase their own property. Renters typically have fewer savings and may earn insufficient salaries to cover rental and essential expenses. Consequently, the combination of rental payments and other financial obligations increases their vulnerability to poverty.

@@

## Relationship betwen Mortgage State and Poverty Status
In the USA in 2019, housing costs accounted for more than 30% of household expenditure, making mortgage status a critical factor in determining a family's risk of falling into poverty [@Statistics_2020]. The percentage of individuals below the poverty threshold, adjusted for housing costs are: $29,080 for a family of four with a mortgage, $24,413 for a family of four without a mortgage, and $29,194 for a family of four renting [@Khattar_Pathak_Schweitzer_Khan_McManus_Jones_Benson_Kelly_2023]. From this, we can assumed that families without mortgages experienced the lowest housing costs, leaving more funds available for other essentials. In contrast, families with mortgages may face burdensome monthly payments, especially if they have chosen a floating interest rate, making them vulnerable to fluctuations in interest rates. Renters also encounter challenges, as rent increases can strain finances and necessitate either accepting higher housing costs or undertaking the time-consuming and costly process of finding new accommodations.

Also, low-income households face challenges with accessing lower mortgage rates and reducing debt burdens through refinancing. Households with low to moderate incomes were less able to take advantage of lower mortgage rates and alleviate their debt burdens through refinancing during the period of exceptionally low rates: merely 24% of mortgages in low-income regions underwent refinancing, in contrast to the 42% in high-income areas [@mortgage]. Therefore, the low-income households have harder experience with repaying their mortgage than higher income households.

These housing-related challenges increases the risk of poverty for affected families, underscoring the importance of addressing housing affordability as a key component of poverty alleviation efforts.

## Relationship betwen Income and Poverty Status
The relationship between a household's income after tax and its poverty status is characterized by a negative correlation, wherein higher income corresponds to a reduced probability of experiencing poverty. This correlation shows the role of financial resources in mitigating economic hardship and facilitating socio-economic well-being. However, in the United States, the cost of living is notably high, low-income households often struggle to meet basic needs, resulting in an unending cycle of poverty. About half of lower-income parents (52%) say they have not had enough money for food or their rent or mortgage [@Braga_2022]. Some 45% of employed parents with lower incomes say they would be very worried about losing pay if they needed to take a day or two off from work to deal with a sick child or because of child care issues. This type of delimma for low-income households makes them harder to pull themselves out of poverty.

In 2019, the poverty guideline for a one-person household was set at $12,490 annually [@ASPE]. So, it is evident that families earning less than $5,000 are definitely living below the poverty line. Conversely, individuals with high incomes are less likely to experience poverty, as their financial resources enable them to comfortably cover living expenses. They will be able to have savings accumulation, further consolidating their financial security and contributing to wealth accumulation over time.
